Parsing Postfix log files

### 2009-02-25 doc/logparser-acronyms.tex
Corrections to the background chapter.

\newglossaryentry{SMTP}{
    type=\acronymtype,
    name={SMTP},
    description={Simple Mail Transfer Protocol},
    first={Simple Mail Transfer Protocol (SMTP)}
}
\newglossaryentry{SMTP glossary}{
    name={Simple Mail Transfer Protocol},
    description={
        SMTP is the protocol that transfers mail from the sender to the
        recipient across the Internet.  It is a simple, human readable,
        plain text protocol, making it quite simple to test and debug
        problems with it.  A detailed description of SMTP is beyond the
        scope of this thesis: the original protocol definition is
        RFC~821~\cite{RFC821}, updated in RFC~2821~\cite{RFC2821};
        introductory guides can be found
        at~\cite{smtp-intro-01,smtp-intro-02}.
    }
}

@@

>>>>

\newglossaryentry{SMTP}{
    type=\acronymtype,
    name={SMTP},
    description={Simple Mail Transfer Protocol},
    first={Simple Mail Transfer Protocol (SMTP)}
}
\newcommand{\SMTPglossaryDescription}[0]{%
    \acronym{SMTP} is the protocol used for transferring mail between the
    sending and receiving \acronym{MTA}\@.  It is a simple, human readable,
    plain text protocol, making it quite simple to test and debug problems
    with it.  A detailed description of SMTP is beyond the scope of this
    thesis: the original protocol definition is in RFC~821~\cite{RFC821},
    updated in RFC~2821~\cite{RFC2821}; introductory guides can be found
    at~\cite{smtp-intro-01,smtp-intro-02}.%
}
\newglossaryentry{SMTP glossary}{
    name={Simple Mail Transfer Protocol},
    description={
        \SMTPglossaryDescription{}
    }
}

@@

\newglossaryentry{SPF}{
    type=\acronymtype,
    name={SPF},
    description={Sender Policy Framework},
    first={Sender Policy Framework (SPF)}
}

### 2009-02-25 doc/logparser-background.tex
Proofread bibliography and made corrections.
Corrections to the background chapter.

This section provides background information helpful in understanding the
remainder of this thesis.  It begins with a discussion of the motivation
underlying the project, followed by a brief introduction to
\acronym{SMTP}\@.  It finishes with a longer introduction to Postfix,
concentrating on anti-spam restrictions and policy servers.

@@

This thesis and the program it describes are part of a larger project to
optimise a mail server's Postfix-based anti-spam restrictions, generate
statistics and graphs, and provide a platform on which new restrictions can
be evaluated to determine if they are beneficial in the fight against spam.
The program parses Postfix log files and populates a database with the data
gleaned from those log files, providing a consistent and simple view of the
log files that future tools can utilise.  The gathered data can be used to
optimise current anti-spam measures, to provide a baseline to test new
anti-spam measures against, or to produce statistics showing how effective
those measures are.

Determining which Postfix restrictions reject the highest number of mails
is a short example of the optimisation possible using data from the
database:

\begin{verbatim}
SELECT name, description, restriction_name, hits_total
    FROM rules
    WHERE postfix_action = 'REJECTED'
    ORDER BY hits_total DESC;
\end{verbatim}

If the database supports sub-selects (where the results of one query are
used in another), percentages can be obtained for the top ten restrictions
(sample output shown in \tableref{Sample output from SQL query showing
percentages}):

@@

\begin{table}[ht]
    \caption{Sample output from SQL query showing percentages}
    \empty{}\label{Sample output from SQL query showing
    percentages}
    \begin{tabular}[]{lrr}
        \tabletopline{}%
        Restriction & Number of hits & Percentage of hits \\
        \tablemiddleline{}%
        \input{build/include-sample-sql-output}
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

\begin{verbatim}
SELECT name, description, restriction_name, hits,
        (hits * 100.0 /
            (SELECT SUM(hits)
                FROM rules
                WHERE postfix_action = 'REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE postfix_action = 'REJECTED'
        AND hits < 100
    ORDER BY hits ASC;
\end{verbatim}

@@

\section{Simple Mail Transfer Protocol}

@@

The \glsfirst{SMTP}, originally defined in \acronym{RFC}~821~\cite{RFC821}
and updated in \acronym{RFC}~2821~\cite{RFC2821}, is used for transferring
mail between the sending and receiving \acronym{MTA}\@.  It is a simple,
human readable, plain text protocol, simplifying testing and debugging
problems with it.  Despite the simplicity of the protocol many virus and
spam sending programs fail to implement it properly, so requiring strict
adherence to the protocol specification is beneficial in protecting against
spam and viruses.\footnote{\label{footnote:rfc760}Originally all mail
servers adhered to the principle of \textit{Be liberal in what you accept,
and conservative in what you send\/} from \acronym{RFC}~760~\cite{rfc760},
but unfortunately that principle was written in a friendlier time.  Given
the deluge of spam that mail servers are subjected to daily, a more
appropriate maxim could be: \textit{Require strict adherence to
\acronym{RFC}~2821; implement the strongest restrictions you can; relax the
restrictions and adherence only when legitimate mail is impeded.\/}  It is
not as friendly, nor as catchy, but it more accurately reflects the current
situation.} A typical \acronym{SMTP} conversation resembles the following
(the lines starting with a three digit number are sent by the server, all
other lines are sent by the client):

@@

This example client is so poorly written that not only does it present the
sender and recipient addresses improperly, it ignores the error messages
returned by the server and carries on regardless.  Many spam and virus
sending programs have serious deficiencies; unfortunately others
(particularly newer programs) were written by competent programmers, or
send mail using well written programs (e.g.\ Postfix or Sendmail on Unix
hosts, Microsoft Outlook on Windows hosts).  Traditionally a mail server
would have done its best to deal with deficient clients, with the intention
of accepting as much mail destined for its users as possible, e.g.\ by
ignoring the absence of a HELO command, or by accepting sender or recipient
addresses that were not enclosed in \texttt{<>}.

A detailed description of \acronym{SMTP} is beyond the scope of this
thesis: introductory guides can be found
at~\cite{smtp-intro-01,smtp-intro-02}, and the definitive references
are~\cite{RFC821,RFC2821}.

@@

Postfix is a \glsfirst{MTA} with the following design aims (in order of
importance): security, flexibility of configuration, scalability, and high
performance.  It features extensive optional anti-spam restrictions,
allowing an administrator to employ those restrictions which they judge
suitable for their site's needs, rather than a fixed set chosen by
Postfix's author.  These restrictions can be selectively applied, combined,
and bypassed on a per-client, per-recipient, or per-sender basis, allowing
different levels of stricture and permissiveness.  Postfix leverages simple
lookup tables to support arbitrarily complicated user-defined sequences of
restrictions and exceptions.  Administrators can supply their own rejection
messages to make it clear to senders exactly why their mail was rejected.
Policy servers (\sectionref{policy servers}) provide a simple way to write
new restrictions without having to modify Postfix's source code.
Unfortunately this flexibility has a cost: complexity in the log files
generated.  Although it is relatively simple in most cases to use standard
Unix text processing utilities to determine the fate of an individual mail,
following the journey a mail takes through Postfix can in some cases be
quite difficult.  For most mails the journey is simple and brief, but the
remainder can have quite complex journeys (see \sectionref{complications}
for details).

Postfix's design follows the Unix philosophy of \textit{``Write programs
that do one thing and do it well''\/}~\cite{unix-philosophy}, and it is
separated into multiple programs that each perform one of the tasks
required of an \acronym{MTA}\@: receive mail, send mail, local delivery of
mail, etc.\ (full details can be found in~\cite{postfix-overview}).  Each
log line contains the name of the Postfix component that produced it, and
this information is used when determining which rules should be used to
parse each log line (see \sectionref{rules in architecture} for details).
Postfix's design is strongly influenced by security concerns: those
components that interact with other hosts are not
privileged,\footnote{Privilege means the power to perform actions that are
limited to the administrator, and not available to ordinary users.} so bugs
in those components will not give an attacker extra privileges; those
components that are privileged do not interact with other hosts, making it
much more difficult for an attacker to exploit any bugs that may exist in
those components.

@@

The administrator can define new restrictions as a list of existing
restrictions, allowing arbitrarily long and complex sequences of lookups,
restrictions, and exceptions.  Postfix tries to protect the administrator
from misconfiguration in as far as is reasonable, e.g.\ the restriction
\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
because the parameter it checks (the hostname given in the HELO command) is
under the control of the remote client.  Despite this, it is possible for
the administrator to make catastrophic mistakes, e.g.\ rejecting all mail
--- the administrator must be cognisant of the effects their configuration
changes will have.  This is similar to one of UNIX's design philosophies:
\textit{``UNIX was not designed to stop its users from doing stupid things,
as that would also stop them from doing clever
things''\/}~\cite{unix-philosophy}.

For further information about Postfix restrictions
see~\cite{smtpd_access_readme,smtpd_per_user_control,policy-servers}.

@@

A policy server~\cite{policy-servers} is an external program consulted by
Postfix to determine the fate of an \acronym{SMTP} command.  The policy
server is given state information by Postfix (sample state information is
shown in \tableref{Example attributes sent to policy servers}) and returns
a result (reject, permit, dunno, a rejection name) as described in
\sectionref{Mixing and matching Postfix restrictions}.  A policy server can
perform more complex checks than those provided by Postfix: a trivial
example is allowing addresses associated with the payroll system to send
mail on the third Tuesday after pay day only, to help prevent problems from
phishing mails using faked sender addresses.\footnote{A phishing mail might
claim that the payroll system had a disastrous disk failure, and until the
server is replaced all salary payments will have to be processed manually,
so please reply to this mail with your name, address, and bank account
details.}

@@

    \caption{Example attributes sent to policy servers}
    \empty{}\label{Example attributes sent to policy servers}

@@

    \item Sender Policy Framework (SPF)~\cite{openspf}.  SPF\label{spf
        introduction} records specify which mail servers are allowed to
        send mail claiming to be from a particular domain.  The intention
        is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter}, and
        joe~jobs~\cite{Wikipedia-joe-job}\glsadd{Joe job}.  There has been
        considerable resistance to the proposal because it breaks or vastly
        complicates some features of \acronym{SMTP}, e.g.\ forwarding mail
        from one domain to another when a user moves.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects a delivery attempt when the triple of \newline{}
        \tab{}\texttt{(sender address, recipient address, remote
        \acronym{IP} address)}\newline{} has not been seen before; on
        second and subsequent delivery attempts from that triple the mail
        will be accepted.  This blocks spam from some senders because
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spam sender, but a
        legitimate mail server must retry deliveries that failed
        temporarily.  Sadly spam senders are using increasingly complex and
        well written programs to distribute spam, frequently using an ISP
        provided \acronym{SMTP} server from a compromised machine on the
        ISP's network.  Greylisting will slowly become less useful, but it
        does block a large percentage of spam mail at the moment; the most
        effective restrictions from the \numberOFlogFILES{} log files used
        when generating the results in \sectionref{Results} are shown in
        \tableref{Summary of rejections}.  The table shows that greylisting
        is worth using at the moment, particularly when you take into
        account its position as the final restriction that a mail must
        overcome in the configuration used on the mail server that
        generated the log files: on that server greylisting only takes
        effect for mails that have passed all other restrictions.  Some
        problems may be encountered when using greylisting: some servers
        fail to retry after a temporary failure, or legitimate mail may be
        delayed, particularly when coming from a pool of servers.

@@

    \item Scoring systems such as Policyd-weight~\cite{policyd-weight}
        perform tests on features of the delivery attempt (e.g.\
        \acronym{IP} address, sender address), incrementing or decrementing
        a score based on the results; if the eventual score is higher than
        a threshold the mail is rejected.  The administrator must manually
        whitelist clients if they are to bypass a Postfix restriction,
        whereas using a threshold that requires a delivery attempt to hit
        several scored restrictions frees the administrator from
        whitelisting clients that fall foul of one binary restriction only.

>>>>

This section provides background information helpful in understanding the
remainder of this thesis.  It begins with a discussion of the motivation
underlying the project, followed by a brief introduction to \acronym{SMTP},
and finishes with a longer introduction to Postfix, concentrating on
anti-spam restrictions and policy servers.

@@

This work is part of a larger project to optimise a mail server's
Postfix-based anti-spam restrictions, generate statistics and graphs, and
provide a platform on which new restrictions can be evaluated to determine
if they are beneficial in the fight against spam.  The program written for
this project, \parsername{}, parses Postfix log files and populates a
database with the data gleaned from those log files, providing a consistent
and simple view of the log files that future tools can utilise.  The
gathered data can be used to optimise current anti-spam measures, to
provide a baseline to test new anti-spam measures against, or to produce
statistics showing how effective those measures are.

A short example of the optimisation possible using data from the database
is determining which Postfix restrictions reject the highest number of
mails:

\begin{verbatim}
SELECT name, description, restriction_name, hits_total
    FROM rules
    WHERE action = 'DELIVERY_REJECTED'
    ORDER BY hits_total DESC;
\end{verbatim}

If the database supports sub-selects (where the results of one query are
used as a parameter in another), percentages can be obtained for the top
ten restrictions (sample output shown in \tableref{Sample output from SQL
query}):

XXX db.full-run.sq3 NEEDS TO BE UPDATED AND EVERYTHING REBUILT\@.

@@

\begin{table}[ht]
    \caption{Sample output from SQL query}
    \empty{}\label{Sample output from SQL query}
    \begin{tabular}[]{lrr}
        \tabletopline{}%
        Restriction & Number of  & Percentage of    \\
                    & rejections & total rejections \\
        \tablemiddleline{}%
        \input{build/include-sample-sql-output}
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

\begin{verbatim}
SELECT name, description, restriction_name, hits,
        (hits * 100.0 /
            (SELECT SUM(hits)
                FROM rules
                WHERE action = 'DELIVERY_REJECTED'
            )
        ) || '%' AS percentage
    FROM rules
    WHERE action = 'DELIVERY_REJECTED'
        AND hits < 100
    ORDER BY hits ASC;
\end{verbatim}

@@

\section{Simple Mail Transfer Protocol (SMTP)}

@@

\SMTPglossaryDescription{} Despite the simplicity of the protocol many
virus and spam sending programs fail to implement it properly, so requiring
strict adherence to the protocol specification is beneficial in protecting
against spam and viruses.\footnote{\label{footnote:rfc760}Originally all
mail servers adhered to the principle of \textit{Be liberal in what you
accept, and conservative in what you send\/} from
\acronym{RFC}~760~\cite{rfc760}, but unfortunately that principle was
written in a friendlier time.  Given the deluge of spam that mail servers
are subjected to daily, a more appropriate maxim could be: \textit{Require
strict adherence to relevant \acronyms{RFC}; implement the strongest
restrictions you can; relax the restrictions and adherence only when
legitimate mail is impeded.\/}  It is neither as friendly nor as catchy,
but it more accurately reflects the current situation.} A typical
\acronym{SMTP} conversation resembles the following (the lines starting
with a three digit number are sent by the server, all other lines are sent
by the client):

@@

This example client is so poorly written that not only does it present the
sender and recipient addresses improperly, it ignores the error messages
returned by the server and carries on regardless.  Many spam and virus
sending programs have serious deficiencies; unfortunately other programs,
particularly newer programs, were written by competent programmers or send
mail using well written programs (e.g.\ Postfix or Sendmail on Unix hosts,
Microsoft Outlook on Windows hosts).  Traditionally a mail server would
have done its best to deal with deficient clients, with the intention of
accepting as much mail sent to its users as possible, e.g.\ by ignoring the
absence of a HELO command, or by accepting sender or recipient addresses
that were not enclosed in \texttt{<>}.

@@

Postfix is a \glsfirst{MTA} with the following design aims (in order of
importance): security, flexibility of configuration, scalability, and high
performance.  It features extensive optional anti-spam restrictions,
allowing an administrator to employ those restrictions which they judge
suitable for their site's needs, rather than a fixed set chosen by
Postfix's author.  These restrictions can be selectively applied, combined,
and bypassed on a per-client, per-recipient, or per-sender basis, allowing
different levels of stricture and permissiveness.  Administrators can
supply their own rejection messages to make it clear to senders exactly why
their mail was rejected.  Policy servers (\sectionref{policy servers})
provide a simple way to write new restrictions without having to modify
Postfix's source code.  Unfortunately this flexibility has a cost:
complexity in the log files generated.  Although it is relatively simple in
most cases to use standard Unix text processing utilities to determine the
fate of an individual mail, in some cases it can be quite difficult.  For
most mails the journey is simple and brief, but the remainder can have
quite complex journeys (see \sectionref{complications} for details).

Postfix's design follows the Unix philosophy of \textit{``Write programs
that do one thing and do it well''\/}~\cite{unix-philosophy}, and it is
separated into multiple components that each perform one of the tasks
required of an \acronym{MTA}\@: receive mail, send mail, local delivery of
mail, etc.;\ full details can be found in~\cite{postfix-overview}.
Postfix's design is strongly influenced by security concerns: those
components that interact with other hosts are not
privileged,\footnote{Privilege means the power to perform actions that are
limited to the administrator, and not available to ordinary users.} so bugs
in those components will not give an attacker extra privileges; those
components that are privileged do not interact with other hosts, making it
much more difficult for an attacker to exploit any bugs that may exist in
those components.

@@

The administrator can define new restrictions as a list of existing
restrictions, allowing arbitrarily long and complex user-defined sequences
of lookups, restrictions, and exceptions.  Postfix tries to protect the
administrator from misconfiguration in as far as is reasonable, e.g.\ the
restriction \texttt{check\_helo\_mx\_access} cannot cause a mail to be
accepted, because the parameter it checks (the hostname given in the HELO
command) is under the control of the remote client.  Despite this, it is
possible for the administrator to make catastrophic mistakes, e.g.\
rejecting all mail --- the administrator must be cognisant of the effects
their configuration changes will have.  This is similar to one of UNIX's
design philosophies: \textit{``UNIX was not designed to stop its users from
doing stupid things, as that would also stop them from doing clever
things''\/}~\cite{unix-philosophy}.

@@

A policy server~\cite{policy-servers} is an external program consulted by
Postfix to determine the fate of an \acronym{SMTP} command.  The policy
server is given state information by Postfix (sample state information is
shown in \tableref{Example attributes sent to policy servers}) and returns
a result (reject, permit, dunno, a rejection name) as described in
\sectionref{Mixing and matching Postfix restrictions}.  A policy server can
perform more complex checks than those provided by Postfix: a trivial
example is allowing addresses associated with the payroll system to send
mail on the third Tuesday after pay day only, to help prevent problems from
phishing mails using faked sender addresses. E.g.\ a phishing mail might
claim that the payroll system had a disastrous disk failure, and until the
server is replaced all salary payments will have to be processed manually,
so please reply to this mail with your name, address, and bank account
details; the criminal can then use any details sent to him to help with
identity theft.

@@

    \caption{Example attributes sent to policy servers, taken
    from~\cite{policy-servers}}
    \empty{}\label{Example attributes sent to policy servers}

@@

    \item \acronym{SPF}~\cite{openspf}.  \acronym{SPF}\label{spf
        introduction} records specify which mail servers are allowed to
        send mail using sender addresses from a particular domain.  The
        intention is to reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter}, and
        Joe~jobs~\glsadd{Joe-job}.  There has been considerable resistance
        to \acronym{SPF} because it breaks or vastly complicates some
        commonly-used features of \acronym{SMTP}, e.g.\ forwarding mail
        from one domain to another when a user moves.

    \item Greylisting~\cite{greylisting} is a technique that temporarily
        rejects a delivery attempt when the triple of \newline{}
        \tab{}\texttt{(sender address, recipient address, remote
        \acronym{IP} address)}\newline{} has not been seen before; on
        second and subsequent delivery attempts from that triple the mail
        will be accepted.  This blocks spam from some senders because
        maintaining a list of failed addresses and retrying after a
        temporary failure is uneconomical for a spam sender, but a
        legitimate mail server must retry deliveries that failed
        temporarily.  Sadly spam senders are using increasingly complex and
        well written programs to distribute spam, frequently using an ISP
        provided \acronym{SMTP} server from a compromised machine on the
        ISP's network.  Greylisting will slowly become less effective as
        spam senders adapt, but it does block a large percentage of spam
        mail at the moment; the most effective restrictions from the
        \numberOFlogFILES{} log files used when generating the results in
        \sectionref{Results} are shown in \tableref{Summary of rejections}.
        The table shows that greylisting is worth using at the moment,
        particularly when you take into account its position as the final
        restriction that a mail must overcome in the configuration used on
        the mail server that generated the log files: on that server
        greylisting only takes effect for mails that have passed all other
        restrictions.  Some problems may be encountered when using
        greylisting: some servers fail to retry after a temporary failure,
        or legitimate mail may be delayed, particularly when coming from a
        pool of servers.

@@

    \item Scoring systems such as Policyd-weight~\cite{policyd-weight}
        perform tests on features of the delivery attempt (e.g.\
        \acronym{IP} address, sender address), incrementing or decrementing
        a score based on the results; if the eventual score is higher than
        a threshold the mail is rejected.  The administrator must manually
        whitelist clients if they are to bypass a Postfix restriction,
        whereas using a threshold that requires a delivery attempt to hit
        several scored restrictions will allow mail from clients that would
        fall foul of one binary restriction only.

        XXX SHOULD I URLS OR BIBLIOGRAPHY ENTRIES\@?  SHOULD I SCRAP ALL
        THE URLS AND MOVE EVERYTHING INTO THE BIBLIOGRAPHY\@?  ASK CARL\@.
